\MakeUppercase{Linear versus centred chromatic numbers}\thanks{This research was partly funded by NSERC.}

### 2022-05-09 lin-vs-cen.tex
Cool

We extend these definitions to vertices of a $k\times k$ pseudogrid $G$ with grid-partition $\mathcal{P}:=\{V(P_\mu):\mu\in\VE(G_k)\}$ as follows. For any $\mu\in\VE(G_k)$, define
\[
   \tilde{B}_r(\mu) := \bigcup_{\nu\in \VE(G_r(\mu))} V(P_\nu)
\]
and, for any $v\in V(G)$, let $\tilde{B}_r(v):=\tilde{B}_r(\mu_v)$ where $P_{\mu_v}$ is the unique part in $\mathcal{P}$ that contains $v$.\footnote{Technically, the notation for $\tilde{B}_r(v)$ should include the partition $\mathcal{P}$, but we omit this since there will never be any ambiguity as to which partition is being used.}
For any $v\in V(G)$, define $\tilde{G}_r(v):=G[\tilde{B}_r(v)]$.\todo{Use $B$, $G$, $\tilde{B}$, and $\tilde{G}$ consistently!}

@@

\begin{lem}\label{doubled_colour_set}
  For every integer $r\ge 1$ there exists an integer $d\in O(r^4)$ such that the following is true, for every integer $k\ge 1$.\newline
  Let $G$ be a $k\times k$ pseudogrid with grid-partition $\mathcal{P}:=\{V(P_\mu):\mu\in\VE(G_k)\}$ and let $\varphi$ be a vertex colouring of $G$ with the property that, for each $A\subseteq\varphi(V(G))$, $|\varphi_{\mathcal{P}}^{-1}(A)\cap\interior_r(G_k)|\ge d|A|$.
  Then there exists $S\subseteq V(\interior_r(G))$ such that
  \begin{compactenum}[(i)]
    \item \label{hits_both} for each $\alpha\in\varphi(V(G))$, $|\varphi^{-1}(\alpha)\cap S|= 2$; and
    \item \label{spread_out} for each $v\in S$, $|\tilde{B}_r(v)\cap S|\le 2$ for each $v\in S$.
  \end{compactenum}
\end{lem}

@@

  We construct $Q$ in two rounds.  In the first round, we start with an initially empty set $Q_1$ and repeat the following for each $\alpha\in\varphi(V(G))$:
  If there exists distinct $\mu_1,\mu_2\in\VE(G_k)$ such that
  \begin{compactenum}[(a)]
    \item $\phi(\mu_1)=\phi(\mu_2)=\alpha$; and\label{hits_both_q}
    \item for each $b\in\{1,2\}$, $\mu_b\not\in \VE(G_r(\mu_{3-b})) \cup \bigcup_{\mu\in Q_1} \VE(G_r(\mu))$,\label{spread_out_q}
  \end{compactenum}
  then we include $\mu_1$ and $\mu_2$ in $Q_1$ and declare the first round an \defin{$\alpha$-success}.  Otherwise, we declare the first round an \defin{$\alpha$-failure}.

@@

  Define a bipartite graph $H$ with parts $L(H):=X:=\varphi(V(G))\setminus \overline{X}$ and $R(H):=Y:=Q_1$.  Let $\pi:=\mu_1,\ldots,\mu_{2|K|}$ be a random permutation of $Y$.  We include an edge $\alpha\mu_i$ in $H$ if and only if
  \[
    \VE(G_r(\mu_i))\setminus\left(\bigcup_{j=1}^{i-1}\VE(G_{2r}(\mu_j))\right)
  \]
  contains some object $\nu$ of colour $\phi(\nu)=\alpha$.  Note the asymmetry here: In the first round, each element $\mu_j$ that takes part in $Q_1$ prevents elements in $\VE(G_r(\mu))$ from also being part of $Q_1$.  In the second round, $\mu_j$ defines a larger box $\VE(G_{2r}(\mu_j))$ that must be avoided when choosing a second element from $\VE(G_r(\mu_i))$, for each $i>j$.

  We now want to use the Local Lemma (\cref{weighted_lovasz}) to show that, with probability greater than zero, $\deg_H(\alpha)\ge 2\vol{r}$ for each $\alpha\in X$.
  For each $\alpha\in X$, let $\Phi_\alpha:=\{\nu\in\VE(G_k):\phi(\nu)=\alpha\}$ and recall that, since $\phi$ comes from \cref{one_colour_per_object}, $|\Phi_\alpha| \ge d$.  Furthermore all but at most $\vol{r}$ elements of $\Phi_\alpha$ are contained in $\bigcup_{\mu\in Y}\VE(G_{r+1}(\mu))$ since otherwise the first round would have been an $\alpha$-success.\todo{Explain more or redefine $\Phi_\alpha$ as $\Phi_\alpha\cap (\bigcup_{\mu\in Y}\VE(G_{r+1}(\mu)))$)}
  % For each $\mu\in Y$, $|\VE(G_{r+1}(\mu))|\in O(r^2)$. Therefore, there are $\Omega(|\Phi_\alpha|/r^2)$ objects $\mu\in Y$ such that $\VE(G_{r+1}(\mu))\cap \Phi_\alpha\neq\emptyset$.

  We begin by showing that, for each $\alpha\in X$, the random variable $\deg_H(\alpha)$ dominates a $\binomial(\ceil{|\Phi_\alpha|/\vol{10r}},1/25)$ random variable, which allows us to establish en exponential inequality for $\Pr(\deg_H(\alpha) < 2\vol{r})$.  (We make no attempt to optimize constants here.)  Suppose that, for some $\mu\in Y$, $B_r(\mu)\cap \Phi_\alpha$ contains at least one element $\nu$. Then, by \cref{packing_lemma}, the set
  \[
     Y_\nu := \{\xi\in Y:\nu\in\VE(G_{2r}(\xi))\}
  \]
  has size at most $25$. Observe that the edge $\alpha\mu$ will appear in $H$ if, in our random permutation $\pi$ of $Y$, $\nu$ appears before any other element of $Y_\nu$.  This happens with probability exactly $1/|Y_\nu|\ge 1/25$. This already enough to establish that $\E(\deg_H(\alpha)) \in \Omega(|\Phi_\alpha|/r^2)$.

  In order to obtain a sufficiently strong concentration result for $\deg_H(\alpha)$ we need to find some independence.  To do this, set $J:=\Phi_\alpha$ and consider the greedy process of repeatedly choosing some $\mu\in Y$ such that $\VE(G_r(\mu))\cap J$ contains at least one element $\nu$ and then set $J:=J-\VE(G_{10r}(\mu))$.\footnote{The constant $10$ is overkill here and is only used for simplicity; $8r+O(1)$ would be sufficient.}  This process continues until $\VE(G_r(\mu))\cap J=\emptyset$ for each $\mu\in Y$. Since $|\VE(G_{10r}(\mu))|\le \vol{10r}$, each step in this process eliminates at most $\vol{10r}$ elements from $J$ so the number, $t$, of iterations is at least $\ceil{|\Phi_\alpha|/\vol{10r}}$

  Let $\{\mu_1',\ldots,\mu_t'\}$ be the subset of $Y$ chosen by this process and let $\{\nu_1,\ldots,\nu_r\}\subseteq\Phi_\alpha$ be the corresponding elements of $\Phi_\alpha$.  The important observation now is that the sets $Y_{\nu_1},\ldots,Y_{\nu_t}$ are disjoint.  Therefore, if we let $U_i$ denote the event ``$\mu_i'$ apppears in $\pi$ before any other element of $Y_{\nu_i}$'' then the events $U_1,\ldots,U_k$ are mutually independent.  Indeed, each $U_i$ depends only on the relative order of $Y_{\nu_i}$ in the  permutation $\pi$.  For each $i\in\{1,\ldots,k\}$, $\Pr(U_i)\ge 1/25$.  Therefore, $\sum_{i=1}^t \mathbbm{1}_{U_i}$ dominates a $\binomial(t,1/25)$ random variable.\footnote{A random variable $X$ dominates a random variable $Y$ if $\Pr(X\ge x)\ge \Pr(Y\ge x)$ for all $x\in\R$.}  Therefore,
  \begin{align}
    \Pr\left(\deg_H(\alpha)< 2\vol{r}\right)
    & \le \Pr\left(\sum_{i=1}^t \mathbbm{1}_{U_i} < 2\vol{r}\right) \notag \\
    & \le \Pr(\binomial(t,1/25)\le 2\vol{r}) \notag \\
    & \le (2t)^{2\vol{r}}(1/25)^{r^2}(1-1/25)^{t-2\vol{r}} \notag \\
    % & \le \exp(6(2r+1)^2\log(2t)+(t-6(2r+1)^2)\log(1-1/25)) \\
    & \le \exp({2\vol{r}}(1+\log(2t))-t/25) \notag \\
    & \le \exp({2\vol{r}}(1+\log(2|\Phi_\alpha|))-|\Phi_\alpha|/25\vol{10r}) \label{probability}
    % & \le \exp(-6(2r+1)^2\log(2t)\log(1-1/?) + t\log(1-1/25)) \\
    % & = \exp(ar^2 - bt)
  \end{align}
  % for fixed positive $a,b>0$.  Recall that we are working in the regime in  $t\in\Omega(|\Phi_\alpha|/r^2)$ and $|\Phi_\alpha|\in\Omega(d/r^2)$ so, for any fixed $C>0$, there exists $d\in\Omega(r^4)$ such that
  % $\Pr(\deg_H(\alpha)< 6(2r+1)^2)\le \exp(-C|\Phi_\alpha|/r^2)$.\todo{This could be done better.}   By the preceding discussion, $\Pr(E_\alpha)\le \exp(-C|\Phi_\alpha|/r^2)$.

>>>>

We extend these definitions to vertices of a $k\times k$ pseudogrid $G$ with grid-partition $\mathcal{P}:=\{V(P_\mu):\mu\in\VE(G_k)\}$ as follows. For any $\mu\in\VE(G_k)$, define
\[
   \tilde{B}_r(\mu) := \bigcup_{\nu\in \VE(G_r(\mu))} V(P_\nu)
\]
and, for any $v\in V(G)$, let $\tilde{B}_r(v):=\tilde{B}_r(\mu_v)$ where $P_{\mu_v}$ is the unique part in $\mathcal{P}$ that contains $v$.\footnote{Technically, the notation for $\tilde{B}_r(v)$ should include the partition $\mathcal{P}$, but we omit this since there will never be any ambiguity as to which partition is being used.}
For any $v\in V(G)$, define $\tilde{G}_r(v):=G[\tilde{B}_r(v)]$.

@@

\begin{lem}\label{doubled_colour_set}
  For every integer $r\ge 1$ there exists an integer $d\in O(r^4)$ such that the following is true, for every integer $k\ge 1$.\newline
  Let $G$ be a $k\times k$ pseudogrid with grid-partition $\mathcal{P}:=\{V(P_\mu):\mu\in\VE(G_k)\}$ and let $\varphi$ be a vertex colouring of $G$ with the property that, for each $A\subseteq\varphi(V(G))$, $|\varphi_{\mathcal{P}}^{-1}(A)\cap\interior_r(G_k)|\ge d|A|$.
  Then there exists $S\subseteq V(\interior_r(G))$ such that
  \begin{compactenum}[(i)]
    \item \label{hits_both} for each $\alpha\in\varphi(V(G))$, $|\varphi^{-1}(\alpha)\cap S|= 2$; and
    \item \label{spread_out} for each $v\in S$, $|\tilde{B}_r(v)\cap S|\le 2$.
  \end{compactenum}
\end{lem}

@@

  We construct $Q$ in two rounds.  In the first round, we start with an initially empty set $Q_1$ and repeat the following for each $\alpha\in\varphi(V(G))$:
  If there exists distinct $\mu_1,\mu_2\in\VE(G_k)$ such that
  \begin{compactenum}[(a)]
    \item $\phi(\mu_1)=\phi(\mu_2)=\alpha$; and\label{hits_both_q}
    \item for each $b\in\{1,2\}$, $\mu_b\not\in \VE(G_{2r+2}(\mu_{3-b})) \cup \bigcup_{\mu\in Q_1} \VE(G_{2r+2}(\mu))$,\label{spread_out_q}
  \end{compactenum}
  then we include $\mu_1$ and $\mu_2$ in $Q_1$ and declare the first round an \defin{$\alpha$-success}.  Otherwise, we declare the first round an \defin{$\alpha$-failure}.

@@

  Define a bipartite graph $H$ with parts $L(H):=X:=\varphi(V(G))\setminus \overline{X}$ and $R(H):=Y:=Q_1$.  Let $\pi:=\mu_1,\ldots,\mu_{2|K|}$ be a random permutation of $Y$.  We include an edge $\alpha\mu_i$ in $H$ if and only if
  \[
    \VE(G_{2r+2}(\mu_i))\setminus\left(\bigcup_{j=1}^{i-1}\VE(G_{2r+2}(\mu_j))\right)
  \]
  contains some object $\nu$ of colour $\phi(\nu)=\alpha$.  Note the asymmetry here: In the first round, each element $\mu_j$ that takes part in $Q_1$ prevents any elements in $\VE(G_{2r+2}(\mu))$ from also being part of $Q_1$.  In the second round, $\VE(G_{2r+2}(\mu_j))$ need only be avoided when choosing a second element from $\VE(G_r(\mu_i))$, for $i>j$.

  We now want to use the Local Lemma (\cref{weighted_lovasz}) to show that, with probability greater than zero, $\deg_H(\alpha)\ge 2\vol{2r+2}$ for each $\alpha\in X$.  For each $\alpha\in X$, let $\Phi^0_\alpha:=\{\nu\in\VE(G_k):\phi(\nu)=\alpha\}$ and recall that, since $\phi$ comes from \cref{one_colour_per_object}, $|\Phi^0_\alpha| \ge d$.  Furthermore all but at most $\vol{2r+2}$ elements of $\Phi^0_\alpha$ are contained in $\bigcup_{\mu\in Y}\VE(G_{2r+2}(\mu))$ since otherwise the first round would have been an $\alpha$-success.  Therefore, the set $\Phi_\alpha:=\Phi^0_\alpha\cap (\bigcup_{\mu\in Y}\VE(G_{2r+2}(\mu)))$ has size $|\Phi_\alpha|\ge d-\vol{2r+2}$.

  % \todo{Continue here...}\todo{Explain more or redefine $\Phi_\alpha$ as $\Phi_\alpha\cap (\bigcup_{\mu\in Y}\VE(G_{r+1}(\mu)))$)}
  % For each $\mu\in Y$, $|\VE(G_{r+1}(\mu))|\in O(r^2)$. Therefore, there are $\Omega(|\Phi_\alpha|/r^2)$ objects $\mu\in Y$ such that $\VE(G_{r+1}(\mu))\cap \Phi_\alpha\neq\emptyset$.

  We begin by showing that, for each $\alpha\in X$, the random variable $\deg_H(\alpha)$ dominates a $\binomial(\ceil{|\Phi_\alpha|/\vol{5r}},1/4)$ random variable, which allows us to establish en exponential inequality for $\Pr(\deg_H(\alpha) < 2\vol{2r+2})$.  (We make no attempt to optimize constants here.)  Suppose that, for some $\mu\in Y$, $B_{2r+2}(\mu)\cap \Phi_\alpha$ contains at least one element $\nu$. Then, by \cref{packing_lemma}, the set
  \[
     Y_\nu := \{\xi\in Y:\nu\in\VE(G_{2r+2}(\xi))\}
  \]
  has size at most $4$.\todo{Update \cref{packing_lemma}} Observe that the edge $\alpha\mu$ will appear in $H$ if, in our random permutation $\pi$ of $Y$, $\nu$ appears before any other element of $Y_\nu$.  This happens with probability exactly $1/|Y_\nu|\ge 1/4$. This already enough to establish that $\E(\deg_H(\alpha)) \in \Omega(|\Phi_\alpha|/r^2)$, but in order to obtain a sufficiently strong concentration result for $\deg_H(\alpha)$ we need to find some independence.  To do this, set $J:=\Phi_\alpha$ and consider the greedy process of repeatedly choosing some $\mu\in Y$ such that $\VE(G_{2r+2}(\mu))\cap J$ contains at least one element $\nu$ and then set $J:=J\setminus\VE(G_{5r}(\mu))$.\footnote{The constant $10$ is overkill here and is only used for simplicity; $4r+O(1)$ would be sufficient.}  This process continues until $\VE(G_{2r+2}(\mu))\cap J=\emptyset$ for each $\mu\in Y$. Since $|\VE(G_{5r}(\mu))|\le \vol{5r}$, each step in this process eliminates at most $\vol{5r}$ elements from $J$ so the number, $t$, of iterations is at least $\ceil{|\Phi_\alpha|/\vol{5r}}$

  Let $\{\mu_1',\ldots,\mu_t'\}$ be the subset of $Y$ chosen by this process and let $\{\nu_1,\ldots,\nu_t\}\subseteq\Phi_\alpha$ be the corresponding elements of $\Phi_\alpha$.  The important observation now is that the sets $Y_{\nu_1},\ldots,Y_{\nu_t}$ are disjoint.  Therefore, if we let $U_i$ denote the event ``$\mu_i'$ apppears in $\pi$ before any other element of $Y_{\nu_i}$'' then the events $U_1,\ldots,U_t$ are mutually independent.  Indeed, each $U_i$ depends only on the relative order of $Y_{\nu_i}$ in the  permutation $\pi$.  For each $i\in\{1,\ldots,t\}$, $\Pr(U_i)\ge 1/4$.  Therefore, $\deg_H(\alpha)\ge \sum_{i=1}^t \mathbbm{1}_{U_i}$ dominates a $\binomial(t,1/4)$ random variable.\footnote{A random variable $X$ dominates a random variable $Y$ if $\Pr(X\ge x)\ge \Pr(Y\ge x)$ for all $x\in\R$.}  Therefore,
  \begin{align}
    \Pr\left(\deg_H(\alpha)< 2\vol{2r+2}\right)
    & \le \Pr\left(\sum_{i=1}^t \mathbbm{1}_{U_i} < 2\vol{2r+2}\right) \notag \\
    & \le \Pr(\binomial(t,1/4)\le 2\vol{2r+2}) \notag \\
    & = \sum_{x=0}^{2\vol{2r+2}-1} \binom{t}{x}\cdot (1/4)^x \cdot (3/4)^{t-x} \\
    & \le (2t)^{2\vol{2r+2}}(1/4)^{2\vol{2r+2}}(3/4)^{t-2\vol{2r+2}} \notag \\
    % & \le \exp(6(2r+1)^2\log(2t)+(t-6(2r+1)^2)\log(1-1/25)) \\
    & \le \exp({2\vol{2r+2}}(1+\log(2t))-t/4) \notag \\
    & \le \exp({2\vol{2r+2}}(1+\log(2|\Phi_\alpha|))-|\Phi_\alpha|/4\vol{5r}) \label{probability}
    % & \le \exp(-6(2r+1)^2\log(2t)\log(1-1/?) + t\log(1-1/25)) \\
    % & = \exp(ar^2 - bt)
  \end{align}
  % for fixed positive $a,b>0$.  Recall that we are working in the regime in  $t\in\Omega(|\Phi_\alpha|/r^2)$ and $|\Phi_\alpha|\in\Omega(d/r^2)$ so, for any fixed $C>0$, there exists $d\in\Omega(r^4)$ such that
  % $\Pr(\deg_H(\alpha)< 6(2r+1)^2)\le \exp(-C|\Phi_\alpha|/r^2)$.\todo{This could be done better.}   By the preceding discussion, $\Pr(E_\alpha)\le \exp(-C|\Phi_\alpha|/r^2)$.
